---
title: 'Chapter 2: Mixed-Effects Models for Factorial Designs'
author: "Keith Lohse, PhD, PStat"
date: '2021-06-09'
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

For these examples, we are going to work with a fictional data that that has two 
fully nested (i.e., between-subjects) and two fully crossed (i.e., within-subject)
factors. We have two hypothetical groups of older adults (OA, >65 y) and younger 
adults (YA, <35 y). Within each of those groups, half of the participants were
in the control group (C) and half were in the treatment group (T). Each participant 
was also measured on four different trials (1 to 4) in three different conditions 
(A, B, and C). 

First, we will open the relevant R packages that we will use and then we will 
download the data from the web. (If you don't have these packages installed, you 
will need install the packages before running the library() function.) Once the data 
are downloaded we can use the head() function to view the first ten rows of the data.


```{r, setting libraries, results="hide", message = FALSE, warning=FALSE}
library(tidyverse); library(RCurl); library(ez); library(lme4); library(car)

DATA <- read.csv("https://raw.githubusercontent.com/keithlohse/mixed_effects_models/master/data_example.csv",
                 stringsAsFactors = TRUE)

```

``` {r, print data to screen}
head(DATA, 10)

```

For the examples we will walk through, we will ignore different variables at different 
times. For instance, we will want to average across trials to create a data set with one 
observation per condition (which we will call *data_COND*), and we will want to average across conditions
to create a data set with one observation for each trial (which we will call *data_TIME*).

Averaging across trials, here are the first ten rows of *data_COND*. 
```{r, subsetting data by condition, warning=FALSE}
data_COND <- DATA %>% group_by(subID, condition, age_group, group) %>% # Note we ignore trial to average across it
  summarize(speed = mean(speed, na.rm=TRUE)) %>% # I have included na.rm=TRUE even though there are no missing data
  arrange(age_group, subID, condition) # finally, we can resort our data by subIDs within groups
head(data_COND, 10)
```

Averaging across conditions, here are the first ten rows of *data_TIME*. 
``` {r, subsetting data by time, warning=FALSE}
data_TIME <- DATA %>% group_by(subID, time, age_group, group) %>% # Note we ignore condition to average across it
  summarize(speed = mean(speed, na.rm=TRUE)) %>% # I have included na.rm=TRUE even though there are no missing data
  arrange(age_group, subID, time) # finally, we can resort our data by subIDs within groups
head(data_TIME, 10)

```

With these data in place, we can now test all of our different models. We are going
to explore the appropriate mixed-effects regression (MER) models for these different 
situations:

  1. **MER as One-Way Repeated Measures ANOVA**
      * When we have a single crossed factor (i.e., one within-subjects factor) in 
      our mixed-effects model.
  2. **MER as Two-Way Repeated Measures ANOVA**
      * When we have multiple crossed factors (i.e., two within-subjects factors) in 
    our mixed-effects model.
  3. **MER as Mixed-Factorial ANOVA with a Single Within-Subjects Factor**
      * When we have a single crossed factor (i.e., one within-subjects factor) and 
    a single nested factor (i.e., between-subjects factors) in our mixed-effects 
    model.
  4. **MER as Mixed-Factorial ANOVA with Multiple Within-Subjects Factors**
      * When we have multiple crossed factors (i.e., two within-subjects factor) and 
    a single nested factor (i.e., between-subjects factors) in our mixed-effects 
    model.


# 1. MER as One-Way Repeated Measures ANOVA 
## (A Single Crossed Factor)
For this example, we will focus on only the effect of condition, so we will use 
the data_COND dataset to average across different trials. First, let's plot the 
data to get a better sense of what the data look like. 


``` {r plotting the effects of condition, echo=TRUE, fig.align="center"}
ggplot(data_COND, aes(x = condition, y = speed)) +
  geom_point(aes(fill=condition), pch=21, size=2,
             position=position_jitter(w=0.2, h=0))+
  geom_boxplot(aes(fill=condition), col="black", 
               alpha=0.4, width=0.5, outlier.shape = NA)+
  scale_x_discrete(name = "Condition") +
  scale_y_continuous(name = "Speed (m/s)") +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "none")
  
```

## 1.1. As an ANOVA...
To implement a simple one-way repeated measures ANOVA, we have a few options. We 
could directly code our ANOVA using the aov() function in R:
``` {r one way ANOVA the hard way, echo=TRUE, warning=FALSE}
summary(aov(speed ~ condition + Error(subID/condition), data=data_COND))
```

Or we can use the ezANOVA() function from the "ez" package. 
```{r one way ANOVA the ez way, echo=TRUE, warning=FALSE}
ezANOVA(data = data_COND, 
    dv = .(speed),
    wid = .(subID),
    within = .(condition)
)
```
Although these two different functions present the results in slightly different
ways, note that the f-values for the two different omnibus tests match. By either 
method, the f-observed for the main-effect of condition is **F(2,78)=24.46, p<0.001**. 

If you are familiar with issues of contrast versus treatment coding, Type I
versus Type III Sums of Squared Errors, and collinearity, then directly controlling
your own ANOVA using base R is probably a safe bet. 

If you are not familiar with those terms/issues, then the ezANOVA code is probably the best option for you. However, I would strongly encourage you to develop a more detailed understanding of general-linear models before jumping into mixed-effect models. 

By default the aov() function provides a test of your statistical
model using Type I Sums of Squared Errors, whereas the ezANOVA() function provides
a test of your statistical model using Type III Sums of Squared Errors. In a completely
orthogonal design where all factors are statistically independent of each other 
(i.e., there is no collinearity), the Type I and Type III Sums of Squared Errors 
will agree. By default, R also uses treatment coding (or "dummy" coding) for categorical
factors rather than orthogonal contrast codes. The omnibus F-tests that we see in 
the ANOVA output will be the same regardless of the types of codes used, but if 
we dig into individual regression coefficients, it is important to remember how
these variables were coded so that we can interpret them correctly.


## 1.2. Getting the same result with a mixed-effect model...
Because we have a single within-subject factor, we will  need to add a 
random-effect of subject to account for individual differences between subjects.
By partitioning the between-subjects variance out of our model, we can fairly test
the effect of *condition*, because our residuals will now be independent of each
other. 
```{r one way ANOVA in a mixed effects model, echo=TRUE, warning=FALSE}
# First we will define our model
mod1 <- lmer(speed ~ 
               # Fixed Effects:
               condition + 
               # Random Effects: 
               (1|subID), 
             # Define the data: 
             data=data_COND, REML = TRUE)

# We can then get the ANOVA results for our model:
anova(mod1)
```

Most critically, note that the F-value, **F(2,78)=24.46, p<0.001** is identical in 
both the RM ANOVA and in the mixed-effects regression model.

If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects
and residuals. 

```{r}
summary(mod1)
```







# 2. MER as Two-Way Repeated Measures ANOVA 
## (Multiple Crossed Factors)
For this example, we will be analyzing both Time and Condition, so we will use 
our full data frame, *DATA*. First, let's plot the data to get a better 
sense of what our data look like.

``` {r plotting two repeated measures, fig.align="center"}
DATA$time <- factor(DATA$time)

ggplot(DATA, aes(x = time, y = speed)) +
  geom_point(aes(fill=condition), pch=21, size=2,
             position=position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1))+
  geom_boxplot(aes(fill=condition), col="black", 
               alpha=0.4, width=0.5, outlier.shape = NA)+
  labs(fill="Condition")+
  scale_x_discrete(name = "Trial") +
  scale_y_continuous(name = "Speed (m/s)", limits = c(0,3)) +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16, color="black"), 
        legend.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "bottom")
``` 


## 2.1. As an ANOVA...
Before we run our models, we want to convert trial to a **factor** so that our model
is treating time categorically rather than continuously. (In later modules, we 
will discuss how to mix continuous and categorical factors.) Additionally, remember
that we are now using our larger data set *DATA* rather than our aggregated data
set *data_COND*. To implement a two-way repeated measures ANOVA, we have the same 
options as before. We can directly code our ANOVA using the aov() function in R:
``` {r two way ANOVA the hard way, echo=TRUE}
DATA$time <- factor(DATA$time)
summary(aov(speed ~ condition*time + Error(subID/(condition*time)), data=DATA))
```

Or we can use the ezANOVA() function from the "ez" package. 
```{r two way ANOVA the ez way, echo=TRUE}
ezANOVA(data = DATA, 
    dv = .(speed),
    wid = .(subID),
    within = .(time, condition)
)
```
Again, regardless of the coding approach that you use, these two different functions 
produce the same f-observed for the main-effects of time, **F(3,117)=105.95, p<0.001**,
and Condition, **F(2,78)=24.46, p<0.001**, and the Time x Condition interaction, **F(6,234)=13.53, p<0.001**. 

 


## 2.2. Two-way RM ANOVA as a mixed-effect model...
Because we now have two crossed factors, we need to account not only for the fact 
that we have multiple observations for each participant, but we have multiple 
observations for each person at each level of each factor. 

For instance, for the effect of Condition, we actually have Condition effects at Trial 1, Trial 2, Trial 3, and Trial 4. 
The converse is also true for the effect of Time, we have four different trials in
Condition A, Condition B, and Condition C. Thus, in order to appropriately 
account for the statistical dependencies in our data, we need to add random-effects
of "time:subID" and "condition:subID" to the model. 

The colon operator (":") means that we are crossing or multiplying these factors. 
That is, if we have subject ID's A, B, and C and Trials 1, 2, 3, and 4, then we end up with 
A1, A2, A3, A4, B1, B2, B3, B4, etc. 

For more information on how random-effects are specified and what they mean in R,
I recommend looking at this discussion on Stack Exchange: https://stats.stackexchange.com/questions/228800/crossed-vs-nested-random-effects-how-do-they-differ-and-how-are-they-specified 


```{r two way another in a mixed effects model, echo=TRUE, warning=FALSE}
# First we will define our model
mod2 <- lmer(speed ~ 
               # Fixed Effects:
               time*condition + 
               # Random Effects
               (1|subID)+ (1|time:subID) + (1|condition:subID), 
               # Define your data, 
             data=DATA, REML=TRUE)

# We can then get the ANOVA results for our model:
anova(mod2)
```

Again, for our purposes, the critical thing to note is that the F-values for the 
main-effects and interactions are the same between our RM ANOVAs and the mixed-effects
model. Without delving into the mathematical details, this is a good demonstration 
that the appropriate random-effects our regression model make it analogous to the
factorial ANOVA. This allows us to capitalize on the benefits of mixed-effects regression for
designs that we would normally analyze using factorial ANOVA. 


If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects
and residuals. 

```{r}
summary(mod2)
```





# 3. MER as Mixed-Factorial ANOVA with One Crossed Factor 
## (2 (Age) x 3 (Condition) ANOVA Example
For this example, we will now consider both the between-subject factor of age, 
as well as the within-subject factor of condition. Do to this, we will 
average over the different trials (1, 2, 3, and 4) to get one observation in each 
condition. This design is sometimes referred as a "mixed-factorial" design, because
we have a mix of between-subjects and within-subject factors. Depending 
on your background, you might be more familiar with this as a "split plot" design. 
Split plot comes from agronomy (where a lot of statistics where developed!) and 
refers to the fact that some plots are assigned to different levels of Factor A (e.g., Treatment
versus Control), but within each plot there is a second level of randomization to 
different levels of Factor B (e.g., Conditions A, B, or C). Both terms describe 
the same thing but differ in their unit of analysis. In psychology, a single person 
is usually the experimental unit (hence "within-subject" variables) whereas in
agronomy or biology a physical area might be the unit of analysis (hence "split-plot"
variables). The more general way to talk about these terms is as **nested-** versus 
**crossed-**factors. 

* For fully **nested** factors, an experimental unit is represented in only
one of the factors (e.g., a person can only be in the treatment group or the 
control group). 
* For fully **crossed** factors, an experimental unit is represented at
all levels of the factor (e.g., e.g., a person is tested in conditions A, B, and C).

In our example, participants are nested within age-group, because each person
is represented at only one level of that factor (i.e., you are only a younger adult or older adult). 
However, Condition is a crossed factor, because each person is represented at all levels of Condition 
(i.e., each person was measured in all three conditions). We can see this more clearly if we plot all of our data.

``` {r plotting crossed and nested factors, warning=FALSE, fig.align="center"}
ggplot(data_COND, aes(x = condition, y = speed)) +
  geom_point(aes(fill=age_group), pch=21, size=2,
             position=position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1))+
  geom_boxplot(aes(fill=age_group), col="black",
               alpha=0.4, width=0.5, outlier.shape=NA)+
  labs(fill="Age Group")+
  scale_x_discrete(name = "Condition") +
  scale_y_continuous(name = "Speed (m/s)") +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16, color="black"), 
        legend.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "bottom")
``` 


## 3.1. As an ANOVA...
For this mixed factorial ANOVA, we have one factor of Condition that varies 
within-subjects, but we also have a factor of Age Group that 
varies between subjects. As before, we can directly code this into our analysis
of variance using the aov() function or using the ezANOVA() function from the "ez" 
package. 
``` {r mixed factorial ANOVA the hard way, echo=TRUE}
summary(aov(speed ~ age_group*condition + Error(subID/condition), data=data_COND))
```

Or we can use the ezANOVA() function from the "ez" package. 
```{r mixed factorial ANOVA the ez way, echo=TRUE}
ezANOVA(data = data_COND, 
    dv = .(speed),
    wid = .(subID),
    within = .(condition),
    between = .(age_group)
)
```

The ezANOVA() function provides us with more detail, for instance automatically
providing Mauchly's Test for Sphericity and both the Greenhouse-Geisser and Hyunh-Feldt
corrections in the event that the sphericity assumption is violated. Most importantly, however,
the outputs of these two functions agree when we look at the f-statistics for all
main-effects and interactions when sphericity is assumed.

 


## 3.2. Mixed Factorial ANOVA as a mixed-effect model...
For this mixed-factorial design, we need to account for the fact that we have
multiple observations coming from each person, so we will add a random-effect of 
"subID". After accounting for this statistical dependence in our data, we can now 
fairly test the effects of Age Group, and Condition with residuals that are
independent of each other. 

```{r mixed factorial ANOVA in a mixed effects model, echo=TRUE}
# First we will define our model
mod3 <- lmer(speed ~ 
               # Fixed Effects:
               age_group*condition + 
               # Random Effects
               (1|subID), 
               # Define your data, 
             data=data_COND, REML=TRUE)

# We can then get the ANOVA results for our model:
anova(mod3)
```

Critically, note that the f-statistics and degrees of freedom all match our ANOVA outputs (when sphericity is assumed), 
with a main-effect of Age Group, F(1,38)=47.18, Condition, F(2,76)=27.27, and the 
Age Group x Condition interaction, F(2,76)=5.48.

If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects,
and residuals. 

```{r}
summary(mod3)
```



# 4. MER as Mixed-Factorial ANOVA with Multiple Crossed Factors 
## 2 (Age Group) x 3 (Condition) x 4 (Trial) ANOVA
Next, we will consider the more complicated example of our fully factorial design
with a nested factor of Age-Group, and crossed factors of Condition and
Time.

``` {r plotting multiple crossed and nested factors, fig.align="center"}
DATA$time <- factor(DATA$time)

ggplot(DATA, aes(x = time, y = speed)) +
  geom_point(aes(fill=age_group), size=2, shape=21,
             position=position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1))+
  geom_boxplot(aes(fill=age_group), col="black", 
               alpha=0.4, width=0.5, outlier.shape = NA)+
  labs(fill="Age Group")+
  facet_wrap(~condition)+
  scale_x_discrete(name = "Trial") +
  scale_y_continuous(name = "Speed (m/s)", limits = c(0,3)) +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16, color="black"), 
        legend.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "bottom")
``` 


## 4.1. As an ANOVA...
For this mixed factorial ANOVA, we have two factors that vary within subjects, 
Condition and Time, and we have one factor that varies between subjects, 
Age Group. As before, we can directly code this into our analysis
of variance using the aov() function or using the ezANOVA() function from the "ez" 
package. 
``` {r multiway factorial ANOVA the hard way, echo=TRUE, warning=FALSE}
summary(aov(speed ~ age_group*condition*time + Error(subID/(condition*time)), data=DATA))
```

```{r multiway factorial ANOVA the ez way, echo=TRUE, warning=FALSE}
ezANOVA(data = DATA, 
    dv = .(speed),
    wid = .(subID),
    within = .(condition, time),
    between = .(age_group)
)
```



## 4.2. Multi-Way Mixed Factorial ANOVA as a mixed-effect model...
For this mixed-factorial design, we need to account for the fact that we have
multiple observations coming from each person, but we also need to acocunt for the 
fact that we multiple observations for each Condition and on each Trial. In order
to account for this dependence in our data, we need to include random-effects
of subject, subject:condition, and subject:time. Adding these random-effects to our 
model will make our mixed-effects model statistically equivalent to the mixed-factorial
ANOVAs that we ran above.

```{r multiway factorial ANOVA in a mixed effects model, echo=TRUE}
# First we will define our model
mod4 <- lmer(speed ~ 
               # Fixed Effects:
               age_group*condition*time + 
               # Random Effects
               (1|subID)+(1|condition:subID)+(1|time:subID), 
               # Define your data, 
             data=DATA, REML=TRUE)

# We can then get the ANOVA results for our model:
anova(mod4)
```

Note that the f-statistics and the degrees of freedom match what we got from the 
ANOVA tables above (when sphericity is assumed). There are a lot of effects so I 
wont list them all, but note the main-effects of Age Group, F(1,38)=47.18, 
Condition, F(2,76)=27.27, and Time, F(3,114)=113.88.

Readers familiar with repeated measures ANOVA wont be surprised to see that that the 
output of the aov() function partitions the effects into different sections. The 
between-subject effect of Age Group is evaluated using residual error at the level of 
"Error:subID". The effects of Condition and Age Group x Condition are evaluated 
using residuals of "Error: subID:condition". This is precisely what we are doing
when we include the three different random-intercepts in the mixed-effects regression.
The $(1|subID)$ partitions out the average individual differences between subjects, 
$(1|condition:subID)$ and $(1|time:subID)$ partitions out the variance at the different
levels of our repeated measures.

If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects
and residuals. 

```{r}
summary(mod4)
```


