---
title: 'Section 7: Generalized Linear Mixed Models (GLMMs) for Non-Normal Data'
author: "Keith Lohse, PhD, PStat"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

Every outcome we have modeled so far in this repository -- gait speed, functional
independence, response time -- has been a continuous measure that we were comfortable treating
as approximately normally distributed. Not every outcome in rehabilitation or movement science
looks like that. Did the participant fall or not (binary)? How many falls did they have in a
month (a count)? What proportion of trials did they complete successfully (a proportion)? For
outcomes like these, a standard linear mixed model (LMM) runs into three related problems:

1. The residuals are structurally non-normal (a binary outcome can only have two possible
   residual values at any given predicted probability).
2. The variance of the outcome is mathematically tied to its mean (for a binomial outcome,
   $\text{Var}(Y) = np(1-p)$; for a Poisson-distributed count, $\text{Var}(Y) = \mu$), so
   "homogeneity of variance" isn't a reasonable assumption.
3. A linear model's predictions are unbounded, so it will happily predict a probability of
   1.3 or a count of $-4$, both of which are impossible. Certainly we might allow for some 
   flexibility here - "interval-like" data - but at a certain point we are better served
   by changing gears and adopting a new model. 

**Generalized Linear Mixed Models (GLMMs)** solve all three problems the same way a standard
generalized linear model (GLM) does: by combining a **link function** $g(\cdot)$ 
that maps the (bounded) mean of the outcome onto an (unbounded) linear predictor, 
with a probability distribution from the exponential family whose variance is 
allowed to depend on its mean in whatever way that distribution requires:
$$g(\mu_{ij}) = \mathbf{X}_{ij} \boldsymbol{\beta} + \mathbf{Z}_{ij} \mathbf{b}_i$$
where $\mu_{ij} = \mathbb{E}[Y_{ij} \mid \mathbf{b}_i]$ is the conditional mean given the
random effects. Everything you already know about specifying fixed and random effects from
Sections 1-4 carries over unchanged; what's new is the choice of link function and
distribution family.

## 7.1. Why `glmmTMB` Instead of `lme4::glmer`?

`lme4::glmer()` is a perfectly good tool for standard binomial and Poisson GLMMs, and if that
is all you need, there is no reason to add another package to your workflow. We use
`glmmTMB` in this section because rehabilitation and clinical outcome data very often need
three things `glmer()` doesn't provide:

* **Overdispersion-aware count families**: native negative binomial distributions
  (`nbinom1` and `nbinom2`), for count data whose variance grows faster than its mean --
  which happen for things like symptom counts, fall counts, or hospital visits.
* **Zero inflation**: a separate model (`ziformula = ~ ...`) for "structural" zeros that
  arise from a different process than the count process itself (e.g., athletes who missed 0 days
  due to injury are probably those that *never got injured*, so 0 versus 1 is a *qualitative* difference,
  whereas 1 versus 2 is a quantitative difference).
* **Modeling the dispersion itself**: `dispformula = ~ ...` lets the residual variance change
  systematically across groups or covariates, similar in spirit to the `varIdent()` structures
  we saw for continuous outcomes in Section 6.

`glmmTMB` is built on Template Model Builder (C++ with automatic differentiation), which also
tends to make it faster than `glmer()` for complex random-effects structures, though for the
simple examples in this section speed is not the deciding factor.

---

## 2. Example 1: A Binary Logistic GLMM

Consider a hypothetical between-subjects experiment: 40 participants were randomized to
either a Control or a Treatment condition, and each participant completed 20 trials of a
task, with a binary success/failure outcome on each trial.

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(glmmTMB)
library(emmeans)
library(performance)
```

```{r simulate-binary-data}
set.seed(101)
n_sub <- 40
n_trials <- 20

df_bin <- expand.grid(
  subID = factor(1:n_sub),
  trial = 1:n_trials
) %>%
  mutate(
    condition = factor(if_else(as.numeric(subID) <= n_sub / 2, "Control", "Treatment"))
  )

sub_intercepts_logit <- rnorm(n_sub, mean = 0, sd = 1.2)

df_bin <- df_bin %>%
  mutate(
    logit_p = -0.5 + 1.2 * (condition == "Treatment") + sub_intercepts_logit[subID],
    prob    = 1 / (1 + exp(-logit_p)),
    success = rbinom(n(), size = 1, prob = prob)
  )
```

```{r fit-binary-glmm}
mod_bin_tmb <- glmmTMB(
  success ~ condition + (1 | subID),
  data   = df_bin,
  family = binomial(link = "logit")
)

summary(mod_bin_tmb)
```

The fixed-effect coefficients here are on the **logit (log-odds) scale**, which is not a
natural scale for most readers to interpret directly. Just as we used `emmeans` in Section 5
to get model-implied means rather than raw coefficients, we can ask `emmeans` to
back-transform these estimates onto the probability scale:

```{r emm-binary}
emm_bin <- emmeans(mod_bin_tmb, ~ condition, type = "response")
emm_bin
```

## 7.3. Example 2: Count Data -- Poisson, Negative Binomial, and Zero-Inflation

Count outcomes bring a second, distinct problem: **overdispersion**. A Poisson distribution
assumes the variance exactly equals the mean, but many real count outcomes (symptom counts,
number of falls, number of relapses) are considerably more variable than that. On top of
overdispersion, clinical count data often has an excess of "structural" zeros -- patients for
whom the count is always going to be zero, for reasons unrelated to the count process for
everyone else.

```{r simulate-count-data}
set.seed(2026)
n_pts <- 50
n_obs <- 6

df_count <- expand.grid(
  patientID = factor(1:n_pts),
  visit     = 1:n_obs
) %>%
  mutate(
    # As above: assigning group by patientID (not by row) keeps this a genuine
    # between-subjects factor, with every patient in one arm for all of their visits.
    group = factor(if_else(as.numeric(patientID) <= n_pts / 2, "Placebo", "Active"))
  )

# Simulate overdispersed counts (via a negative binomial generating process) with an
# additional 30% chance of a "structural" zero layered on top.
p_zero <- 0.30
df_count <- df_count %>%
  mutate(
    is_structural_zero = rbinom(n(), 1, p_zero),
    lambda = exp(1.5 - 0.4 * visit + 0.6 * (group == "Active") + rnorm(n_pts, 0, 0.4)[patientID]),
    raw_count = rnbinom(n(), size = 1.5, mu = lambda),
    symptom_count = if_else(is_structural_zero == 1, 0L, raw_count)
  )
```

### 7.3.1 Start Simple, and Let Diagnostics Tell You When To Upgrade

It's tempting to jump straight to the most flexible model (negative binomial with zero
inflation), but a more feasible workflow is to start with the simplest plausible model, check
whether it's adequate, and only add complexity where the diagnostics say you need it:

```{r fit-poisson}
mod_poisson <- glmmTMB(
  symptom_count ~ group * visit + (1 | patientID),
  data   = df_count,
  family = poisson(link = "log")
)

check_overdispersion(mod_poisson)
```

The Poisson model's own diagnostic tells us clearly that we have a problem: a dispersion
ratio well above 1 means the data are considerably more variable than a Poisson distribution
allows for, and the test is significant. This is exactly the situation `nbinom2` was built
for, since the negative binomial's variance function ($\text{Var}(Y) = \mu + \mu^2/\theta$)
includes a free parameter ($\theta$) to absorb exactly this kind of extra variability:

```{r fit-negbin}
mod_negbin <- glmmTMB(
  symptom_count ~ group * visit + (1 | patientID),
  data   = df_count,
  family = nbinom2(link = "log")
)

check_overdispersion(mod_negbin)
```

Once we let the model estimate its own dispersion parameter, the overdispersion problem
disappears -- which is reassuring, but is also exactly what we should expect, since the
negative binomial model now has the flexibility to match however much extra variance the data
actually contain. Finally, we can add a zero-inflation component to explicitly model the
excess structural zeros:

```{r fit-zinb}
mod_zinb <- glmmTMB(
  symptom_count ~ group * visit + (1 | patientID),
  ziformula = ~ 1,  # a single, constant zero-inflation probability across all observations
  data      = df_count,
  family    = nbinom2(link = "log")
)
```

### 7.3.2 Comparing the Three Models

```{r compare-count-models}
anova(mod_poisson, mod_negbin, mod_zinb)
```

**A caveat on this comparison.** Formally, a likelihood-ratio test assumes the simpler model
is nested inside the more complex one *and* that the parameter being added is not sitting on
the boundary of its allowed range under the null hypothesis. Both comparisons here bend that
assumption a little: the Poisson-to-negative-binomial step is really testing whether the
negative binomial's dispersion parameter is at its (boundary) limiting value where the
negative binomial reduces to a Poisson distribution, and the negative-binomial-to-ZINB step is
testing whether a probability parameter is at the boundary value of 0. In both cases, the
$\chi^2$ p-value from a standard LRT is a commonly-used approximation rather than an exact
test -- the same general caveat about testing variance components at a boundary that came up
for singular fits in Section 3.5.3 of this repository. In practice, most applied
researchers still use these comparisons as a practical, ordered guide (do I need overdispersion?
do I need zero-inflation on top of that?) alongside AIC, rather than treating the p-values as
exact.

### 7.3.3 Marginal Predictions on the Response Scale

Just as we back-transformed the binary model's predictions above, we can ask for the
zero-inflated model's predicted symptom counts on their natural (response) scale, broken out
by group and visit:

```{r emm-count, message=FALSE}
emmeans(mod_zinb, ~ group | visit, at = list(visit = 1:6), type = "response")
```

Note the `at = list(visit = 1:6)`. Because `visit` is a *continuous* predictor in this model
(not a factor), asking for `~ group | visit` without also specifying `at` will only evaluate
the model at `visit`'s mean (here, 3.5) rather than at each actual visit -- giving you one
combined estimate rather than the visit-by-visit trajectory you're probably after. Whenever
you condition on (`|`) a continuous variable, decide explicitly which values of it you want
predictions at.

## Summary

* GLMMs extend everything you already know about specifying fixed and random effects to
  outcomes that are binary, count, or otherwise non-normally distributed, via a link function
  and an appropriate probability family.
* `glmmTMB` adds negative-binomial, zero-inflated, and dispersion-modeling options on top of
  what `lme4::glmer()` already provides.
* * Check overdispersion on your *simplest* candidate model first; it is the diagnostic that
  tells you whether you need to move from Poisson to negative binomial in the first place,
  not just a box to check after the fact on your final model.
* Likelihood-ratio tests that compare different families, or that test a parameter sitting at
  a boundary value (e.g., "is there any zero-inflation at all?"), should be treated as useful
  approximations rather than exact tests.

## References
Brooks, M. E., Kristensen, K., van Benthem, K. J., Magnusson, A., Berg, C. W., Nielsen, A.,
Skaug, H. J., Maechler, M., & Bolker, B. M. (2017). glmmTMB Balances Speed and Flexibility
Among Packages for Zero-inflated Generalized Linear Mixed Modeling. *The R Journal, 9*(2),
378-400. https://doi.org/10.32614/RJ-2017-066

Agresti, A. (2015). *Foundations of Linear and Generalized Linear Models*. Wiley.
