---
title: 'Section 6: Modeling Autocorrelation and Heteroscedasticity with `nlme`'
author: "Keith Lohse, PhD, PStat"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

Every model we have fit with `lme4::lmer()` so far in this repository makes one assumption
that we have never actually stated out loud - although it was implicit in Section 1's very first equation.

Recall the general form of a mixed-effects model from Section 1 (Equation 1 of the companion
paper):
$$\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\boldsymbol{\gamma} + \boldsymbol{\epsilon}$$
where $\mathbf{y}$ is the vector of every observation across every subject, $\mathbf{X}\boldsymbol{\beta}$
is the fixed-effects part of the model, $\mathbf{Z}\boldsymbol{\gamma}$ is the random-effects
part, and $\boldsymbol{\epsilon}$ is the vector of leftover residuals.

To talk about correlation *within* a subject over time, it helps to zoom in on one subject at a
time -- restricting Equation 1 to just subject $i$'s own rows:
$$\mathbf{y}_i = \mathbf{X}_i\boldsymbol{\beta} + \mathbf{Z}_i\boldsymbol{\gamma}_i + \boldsymbol{\epsilon}_i$$
This is not a new model. It is the exact same equation, just written one subject's block of
rows at a time -- the same partitioning that Figure 2 of the companion paper shows $\mathbf{Z}$
performing when it "spreads out" the random effects, $\boldsymbol{\gamma}$, into a share for
each subject.

As we've previously discussed, every time we've written out a set of random effects (e.g.,
Section 2's random-slopes model), that subject $i$'s own vector of random deviates follows a
multivariate normal distribution -- for instance, for a random intercept and random slope:
$$\begin{bmatrix}\gamma_{0i}\\ \gamma_{1i}\end{bmatrix} \sim \mathcal{N}\left(\begin{bmatrix}0\\0\end{bmatrix}, \begin{bmatrix}\sigma_0^2 & \sigma_{01}\\ \sigma_{01} & \sigma_1^2\end{bmatrix}\right)$$
We will now give that covariance matrix a name, $\mathbf{G}$, so that $\boldsymbol{\gamma}_i
\sim \mathcal{N}(\mathbf{0}, \mathbf{G})$. Nothing about $\mathbf{G}$ is new here - it is the
same matrix you have already been building throughout Sections 1-4, just represented by a single
letter instead of spelled out in full every time.

What every model so far *has* quietly assumed, without ever writing it down this explicitly,
is the analogous statement for the residuals: $\boldsymbol{\epsilon}_i \sim
\mathcal{N}(\mathbf{0}, \sigma^2\mathbf{I})$ - every residual for subject $i$ has the same
variance $\sigma^2$, and none of them correlate with one another. We can give this residual covariance
matrix a name too, $\mathbf{R}_i$, so that $\boldsymbol{\epsilon}_i \sim \mathcal{N}(\mathbf{0},
\mathbf{R}_i)$. Up until now, $\mathbf{R}_i$ has simply *been* $\sigma^2\mathbf{I}$, we just
never called it that.

Once both matrices have names, the total variance of subject $i$'s outcomes follows directly.
$\mathbf{X}_i\boldsymbol{\beta}$ is a fixed quantity, not a random one, so it contributes
nothing to the variance; and because $\boldsymbol{\gamma}_i$ and $\boldsymbol{\epsilon}_i$ are
assumed independent of one another,
$$\text{Var}(\mathbf{y}_i) = \text{Var}(\mathbf{Z}_i\boldsymbol{\gamma}_i + \boldsymbol{\epsilon}_i)
= \mathbf{Z}_i\,\text{Var}(\boldsymbol{\gamma}_i)\,\mathbf{Z}_i^T + \text{Var}(\boldsymbol{\epsilon}_i)
= \mathbf{Z}_i \mathbf{G} \mathbf{Z}_i^T + \mathbf{R}_i$$

Thus far, we have principally discussed how to deal with statistical dependence through what is called the **"G-side"** matrix; 
that is, the specification of the **random effects**. However, especially when we have repeated
observations at the lowest level of our data, we might also need to worry about the **"R-side"** matrix; that is, 
are there additional considerations we need to make about the (co)variance of our **residuals** above and beyond our random effects.
Every model in Sections 1-4 fixed $\mathbf{R}_i = \sigma^2\mathbf{I}$ and let $\mathbf{G}$ do all of the work of 
describing dependency within a subject. `lme4` doesn't give you a way to change this -
$\mathbf{R}_i = \sigma^2\mathbf{I}$ is baked into how the package is built, which is part of
why it is so fast even for complicated crossed random-effects structures (Section 8). Critically, 
if we have specified our random effects structure sufficiently, this is also a very reasonable 
assumption. However, treating $\mathbf{R}_i$ as a plain $\sigma^2\mathbf{I}$ may not always 
be sufficient. Specifically, there are two properties of our residuals that we need to 
concern ourselves with:

1. **Autocorrelation.** Observations closer together in time, from the same subject, tend to
   be *more* similar to each other than observations further apart. A person having an
   unusually good day at Week 3 is a better predictor of Week 4 than of Week 40.
2. **Heteroscedasticity.** The variance of the residuals can change systematically over time,
   across conditions, or across groups -- e.g., people's scores might be tightly clustered
   right after admission and much more spread out a year later, as some people have
   recovered fully and others have not.

The `nlme` package (which we already used for the negative-exponential growth model in
Section 2) lets us model both of these directly, by specifying an explicit structure for
$\mathbf{R}_i$ -- on top of whatever $\mathbf{G}$ we've already chosen for the random effects --
instead of accepting the default $\sigma^2\mathbf{I}$.


## 6.1. Common Residual Covariance Structures

* **Compound Symmetry (`corCompSymm`)**: a single correlation $\rho$ between *any* two time
  points from the same subject, regardless of how far apart they are. This is the same
  assumption implicit in a repeated-measures ANOVA when sphericity holds (see Section 3; 
  a random-intercept-only mixed model is mathematically equivalent to compound
  symmetry.
* **Autoregressive Order 1 (`corAR1`)**: correlation decays geometrically with each discrete
  time step, $\text{Corr}(\epsilon_t, \epsilon_{t+k}) = \rho^{|k|}$. This is a natural choice
  when your time points are evenly spaced (Trial 1, 2, 3, 4...) and you expect the
  "good day/bad day" carryover to fade steadily rather than staying constant.
* **Continuous AR1 (`corCAR1`)**: the same idea as `corAR1`, but for *continuously* and (potentially)
  *irregularly* spaced time -- exactly the situation we described in Section 2, where our
  spinal-cord-injury data were "collected on different days for different people," so that
  Month 3 for one participant might really be Day 85 and for another Day 95.
* **Unstructured (`corSymm`)**: no assumed pattern at all -- every pair of time points gets
  its own, freely-estimated correlation. This is the most flexible option but also the most
  informationally "expensive": with $T$ time points it estimates $T(T-1)/2$ separate correlation parameters,
  which can quickly become more than your data can support.

---

## 6.2. A Simulated Example: Recovering a Known AR(1) Process

To see these ideas clearly, we will first simulate data where we know the true answer: 30
subjects, measured at 6 evenly-spaced time points, with residuals that (a) truly follow an
AR(1) process with $\rho = 0.65$ and (b) truly grow more variable over time.

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(nlme)
```

```{r simulate-ar1-data}
set.seed(2026)
n_subjects <- 30
n_timepoints <- 6

df_nlme <- expand.grid(
  subID = factor(1:n_subjects),
  time  = 1:n_timepoints
) %>%
  arrange(subID, time)

sigma_base <- 1.0 # baseline variance
rho_true <- 0.65 # true correlation

# Build each subject's residual trajectory in two steps. First, a properly stationary
# AR(1) base process with constant unit variance throughout -- the sqrt(1 - rho_true^2)
# term on each innovation is essential here; without it, the AR(1) recursion's own
# variance compounds on top of whatever heteroscedasticity we add next, and rho and the
# variance pattern stop being cleanly separable. Second, scale that base process by the
# intended heteroscedasticity pattern, so the growing variance is exactly what we say it
# is, and nothing more.
df_nlme$residual <- unlist(lapply(1:n_subjects, function(i) {
  base <- numeric(n_timepoints)
  base[1] <- rnorm(1, 0, 1)
  for (t in 2:n_timepoints) {
    base[t] <- rho_true * base[t-1] + rnorm(1, 0, sqrt(1 - rho_true^2))
  }
  var_scale <- 1 + 0.2 * (0:(n_timepoints - 1))
  sigma_base * sqrt(var_scale) * base
}))

df_nlme <- df_nlme %>%
  mutate(score = 5.0 + 0.8 * time + residual)
```

## 6.3. Fitting Three Increasingly Realistic Models

We'll fit the same fixed effect of `time` three times, adding one piece of residual structure
at each step, so we can watch model fit improve as our assumptions get closer to how the data
were actually generated.

```{r fit-three-models}
# Model 1: standard random-intercept model. i.i.d. residuals -- our lme4 default assumption.
m1_iid <- lme(
  fixed  = score ~ time,
  random = ~ 1 | subID,
  data   = df_nlme,
  method = "REML"
)

# Model 2: adds an AR(1) residual correlation structure within each subject.
m2_ar1 <- lme(
  fixed       = score ~ time,
  random      = ~ 1 | subID,
  correlation = corAR1(form = ~ time | subID),
  data        = df_nlme,
  method      = "REML"
)

# Model 3: adds a variance function on top of the AR(1) structure, letting the residual
# variance differ freely at each time point (varIdent estimates one scale factor per level
# of its grouping variable -- nlme treats `time` as a grouping factor here even though the
# column itself is numeric, so this correctly gives us one variance parameter per time point).
m3_ar1_var <- lme(
  fixed       = score ~ time,
  random      = ~ 1 | subID,
  correlation = corAR1(form = ~ time | subID),
  weights     = varIdent(form = ~ 1 | time),
  data        = df_nlme,
  method      = "REML"
)
```

```{r compare-models}
anova(m1_iid, m2_ar1, m3_ar1_var)
```

Each model is nested inside the next (Model 2 adds one correlation parameter to Model 1;
Model 3 adds five variance-ratio parameters to Model 2), so the likelihood-ratio tests in this
table are valid, and both are highly significant: allowing for autocorrelation improves the
fit substantially, and allowing the residual variance to grow over time improves it further
still. This matches how we built the data, which of course we rigged in advance. The value
of doing this with simulated data first is exactly that you get to confirm the method
recovers a structure you already know is there before trusting it on real data where you
don't.



## 6.4. Visualizing the Autocorrelation Directly

Beyond checking the numeric value of $\hat\rho$, it is good practice to look at the residual
autocorrelation function (ACF) directly, both before and after adding the `corAR1` term. Keep in 
mind that for this example, we have "baked in" a specific auto-correlation for our simulated data. 
In reality, we don't know what the "true" autocorrelation is, so instead we would be 
looking for any evidence of residual autocorrelation:

```{r acf-plot-1, fig.align="center", fig.width=5, fig.height=4}
# ACF()'s default residuals (resType = "pearson") are NOT adjusted for a model's fitted
# correlation structure, so they will not show whether corAR1 actually worked. Use
# resType = "normalized" instead. See the "Technical notes" here for the same fix in
# another worked example: https://bbolker.github.io/mixedmodels-misc/notes/corr_braindump.html
plot(ACF(m1_iid, maxLag = 5, resType = "normalized"), main = "Model 1 (i.i.d.) Residual ACF")
```

```{r acf-plot-2, fig.align="center", fig.width=5, fig.height=4}
plot(ACF(m2_ar1, maxLag = 5, resType = "normalized"), main = "Model 2 (AR1) Residual ACF")
```


## 6.5. A Real-Data Example: Continuous Time with `corCAR1`

The simulated example above uses evenly-spaced integer time points (1 through 6), which is
exactly the situation `corAR1` was built for. But recall the longitudinal spinal-cord-injury
data from Section 2 (`data_SCI_longitudinal_example.csv`): "these data were all collected on
different days for different people... Month 1 as a time point might be Day 20 for some
people, but Day 30 for others." That is precisely the situation `corCAR1` (continuous-time
AR1) is designed for, and we don't need to simulate anything new to see it in action -- we can
go straight back to real data we've already loaded once before.

It's worth pausing on what Section 2 actually *did* with this exact dependency problem,
because it is a clean illustration of the distinction between G-side and R-side modeling 
that we discussed at the beginning. Section 2 never touched $\mathbf{R}_i$ at all -- 
every one of its models used a plain $\sigma^2\mathbf{I}$ for the
residuals. Instead, it handled the fact that a person's functional-independence scores over
time are not independent entirely through $\mathbf{G}$, by building an increasingly rich
random-effects structure: a random intercept, then a random linear slope for `year.0`, then a
random quadratic term, then a random cubic term -- each addition justified by a significant
likelihood-ratio test ($\chi^2(3)=1046.98$, then $\chi^2(4)=746.46$, then $\chi^2(5)=252.32$).
That is twelve additional variance and covariance parameters, across three model comparisons,
spent on capturing more and more of the within-subject dependency by letting each person's
*own trajectory shape* bend to fit their data. This was a perfectly reasonable strategy --
per Part 6.6 (below), a rich enough $\mathbf{G}$ can produce well-calibrated fixed-effect tests
without ever modeling $\mathbf{R}_i$ explicitly, and Section 2 was, after all, primarily
interested in testing whether AIS Grade predicts the *shape* of that trajectory, which needed
those polynomial fixed effects regardless.

`corCAR1` offers a genuinely different way to attack the same underlying dependency. Instead
of asking $\mathbf{G}$ to become progressively richer, one polynomial term at a time, it
models the persistence directly, in $\mathbf{R}_i$, with a single interpretable parameter:
how strongly today's score predicts a score close in time, decaying continuously the further
apart two assessments are. Let's fit the simplest possible version -- deliberately using only
a linear fixed effect of `year.0` and only a random *intercept*, so that whatever improvement
we see is attributable to $\mathbf{R}_i$ alone, not to a richer $\mathbf{G}$:

```{r real-data-corCAR1, message=FALSE}
DAT2 <- read.csv("https://raw.githubusercontent.com/keithlohse/mixed_effects_models/master/data_SCI_longitudinal_example.csv",
                 stringsAsFactors = TRUE)
DAT2$year.0 <- (DAT2$time - 1) / 12

m_iid_real  <- lme(rasch_FIM ~ year.0, random = ~ 1 | subID, data = DAT2, method = "REML")
m_car1_real <- lme(rasch_FIM ~ year.0, random = ~ 1 | subID,
                    correlation = corCAR1(form = ~ year.0 | subID),
                    data = DAT2, method = "REML")

anova(m_iid_real, m_car1_real)
summary(m_car1_real)$modelStruct$corStruct
```

The improvement in fit here is enormous, and reflective of patterns you'd likely see in real
data -- functional independence really does show strong day-to-day/month-to-month carryover
within a person (someone recovering quickly at Month 3 is very likely to still be 
recovering quickly at Month 4). Notice what it took to get here, however:
Section 2 spent twelve parameters, across an intercept, a slope, a quadratic, and a cubic
random effect, progressively chipping away at within-subject dependency while *also* building
a detailed picture of average recovery curvature. Here, a *single* additional parameter --
$\hat\rho$ -- captures a comparable order of magnitude of improvement (a likelihood-ratio
statistic of 787.84, on just 1 degree of freedom), because it is aimed squarely at the
dependency itself rather than approximating it by modeling trajectory shape.
This is not quite an apples-to-apples comparison -- Section 2's rich $\mathbf{G}$ was doing
double duty, capturing genuine nonlinear curvature *and* individual differences in that
curvature. In contrast, `corCAR1` alone tells us nothing about individual or group level 
trajectories, but it does give us a parsimonious way of dealing with statistical dependency 
in our residuals. 

The two different approaches are not mutually exclusive, either. Nothing stops you from combining them --
fitting Section 2's cubic random-effects growth curve *and* an explicit `corCAR1` term in the
same model, if you have reason to think meaningful residual autocorrelation remains even after
allowing each person their own curve. In practice, with only 40 subjects, that combined model
is asking a great deal of the data at once, and Section 3's discussion of singular fits is a
good reminder to check convergence carefully -- and to consider, per Part 6.6.4, whether your
actual research question (testing a fixed effect vs. interpreting the variance components)
calls for that additional complexity at all -- before trusting it.

This is also a good habit to build for its own sake: whenever this repository already has a
real dataset whose design fits the technique you're learning, it's worth running the new
technique on it, rather than only ever practicing on data you simulated yourself. You already
know quite a lot about this dataset from Section 2 (e.g., the growth curves in Figure/plot
form, the negative-exponential model, the AIS Grade comparisons), and layering `corCAR1` on
top of that existing understanding -- and directly comparing it to the strategy Section 2 already
used -- is a much faster way to build intuition than starting over with an unfamiliar variable
name.

## 6.6. When Does the G-Side/R-Side Distinction Actually Matter?

We now have the tools to model $\mathbf{R}_i$ directly, which raises a sharper question than
"why don't we always use `nlme`": given that $\text{Var}(\mathbf{y}_i) = \mathbf{Z}_i \mathbf{G}
\mathbf{Z}_i^T + \mathbf{R}_i$ simply *adds* the two pieces together, how much does it actually
matter whether a given source of dependency is captured by $\mathbf{G}$ or by $\mathbf{R}_i$?
The honest answer depends heavily on what you are using the model for -- and it is different
for the two things we typically want out of a mixed-effects model: testing fixed effects, and
interpreting the random effects themselves.

### 6.6.1 Fixed-Effect Point Estimates: Robust to the G/R Partition

Every mixed model estimates its fixed effects using a form of generalized least squares:
observations get combined into an estimate of $\boldsymbol{\beta}$, weighted by how much
information each one is assumed to carry, where those weights come directly from the model's
assumed covariance structure, $\mathbf{Z}_i \mathbf{G} \mathbf{Z}_i^T + \mathbf{R}_i$. If that
assumed structure is *somewhat* wrong -- say, because real residual autocorrelation gets
folded into $\mathbf{G}$ instead of being modeled explicitly as $\mathbf{R}_i$ -- the weights
used to combine observations are no longer the *most efficient* possible weights, but they
remain a reasonable, unbiased way to combine the data. This is the same basic robustness
property that keeps ordinary regression coefficients unbiased even under heteroscedasticity:
getting the variance structure wrong costs you precision, and (as the next section shows) can
cost you calibration, but it does not, in general, introduce bias into the point estimate
itself. 

### 6.6.2 Fixed-Effect Standard Errors: Depends on Whether G Is Rich Enough

Standard errors are a different story, because they depend directly on how much
*independent* information the model believes it has. Picture a subject's scores over time
genuinely following an autocorrelated, "streaky" pattern -- a run of a few good days, then a
run of a few bad days -- rather than bouncing around independently from one time point to the
next. A model built only on a random intercept assumes **compound symmetry**: any two of that
subject's time points are equally correlated, no matter how close together or far apart they
are. Such a model has no way to distinguish "four genuinely independent measurements" from
"four measurements drawn from the same short streak." It ends up treating a temporally
clustered, highly-correlated run as if it carried nearly as much information as four
independent looks at the subject's true trend -- which overstates the precision of that trend
estimate and understates its standard error. The risk is highest specifically for fixed
effects that describe *change over time* (like a slope), since that is exactly the kind of
estimate a "streaky" pattern can most easily fool.

A random **slope** changes this picture substantially. Once each subject is allowed their own
individual line -- their own intercept and their own rate of change -- a temporary streak (a
subject who happens to be running high across several adjacent time points) can be partly
absorbed into that subject's own line, instead of being mistaken for extra evidence about the
population-level trend. This does not literally "discover" that an autocorrelated process
generated the data -- the model still has no explicit concept of residual autocorrelation --
but a random slope is flexible enough to approximate a good share of the same pattern (runs of
similar values, within a subject, over nearby time points), which tends to pull the resulting
fixed-effect standard errors back toward being well-calibrated, even without ever touching
$\mathbf{R}_i$ directly.

The practical lesson: **the risk of an anti-conservative fixed-effect standard error is
highest when $\mathbf{G}$ is sparse** -- often just a random intercept -- **and the true
dependency has real structure over time** (autocorrelation, or variance that changes with
time). A random-effects structure with slopes that track the shape of your growth curves,
exactly as Sections 2 and 4 already recommend, substantially closes that gap, because it gives
the model a flexible, subject-specific way to absorb exactly the kind of statistical dependence 
that we would otherwise need to model with a more complex $\mathbf{R}_i$.

### 6.6.3 Random-Effect Variance Components: Where the Partition Matters Much More

Unliketing estimating/testing fixed effects, estimating the variance of random effects
is a different matter entirely, because they *are* the answer to
"how is the total variance divided up" -- and that is precisely the question a G/R
misattribution gets wrong. Picture the total variance
in your outcome as a pie, correctly sized: a model that folds real residual autocorrelation
into $\mathbf{G}$ instead of modeling it explicitly as $\mathbf{R}_i$ is not making the pie
the wrong size -- it is cutting the *same* pie along the wrong lines. Some of what should
properly be described as "unexplained, time-varying noise within a subject" ($\mathbf{R}_i$)
gets relabeled as "this subject's true, stable individual difference" ($\mathbf{G}$), which
typically inflates the estimated between-subject (intercept) variance.

If your model also includes a random slope, the story gets one layer more complicated: some of
that same misattributed variance can get pulled specifically into the *slope* variance instead
of the intercept variance, since a random slope is one of the more flexible tools the model has
available for approximating a "streaky," time-varying pattern. That can make it
look as though people genuinely differ in their rate of change, when a meaningful share of
that apparent variability is actually unmodeled residual dependency, misattributed to the wrong
place in $\mathbf{G}$.

This is the asymmetry worth remembering: **a fixed-effect hypothesis test can come out
well-calibrated using nothing but a rich $\mathbf{G}$, because it only cares about the
*total* variance behind the estimate; but if your actual research question is about the
variance components themselves** -- how much people differ in level, or in trajectory, a
reasonable and clinically meaningful question in rehabilitation research (Sections 2 and 8) --
**the G/R partition matters much more**, because a model can pass a well-calibrated 
fixed-effect test while still giving a distorted answer to random effects.

### 6.6.4 So, When Should You Still Reach for `nlme`?

Putting this together, `nlme`'s explicit $\mathbf{R}_i$ modeling is most worth reaching for
when:

* **The variance components are your actual research question** -- you want to report or
  interpret how much subjects differ in level or trajectory, not just test a fixed effect,
  and you need those estimates to reflect real between-subject differences rather than
  unmodeled residual dependency.
* **Your time structure doesn't fit neatly into a low-order polynomial random-effects
  structure** -- e.g., truly irregular, continuous-time data like Section 2's SCI dataset,
  where `corCAR1` handles the actual observation times directly, rather than approximating
  them through a random slope (and, per Section 2's own warning, high-order polynomial random
  effects bring their own collinearity problems as you add more terms).
* **You have a genuinely sparse $\mathbf{G}$** (often just a random intercept) and reason to
  suspect real autocorrelation or heteroscedasticity beyond what compound symmetry can
  capture -- exactly the situation described in Part 7.2 above, where a sparse $\mathbf{G}$
  produced overconfident, anti-conservative inference about a time trend.
* **The dependency structure is itself the scientific finding** -- e.g., "does variability
  grow over time" is a claim about $\mathbf{R}_i$, and answering it well requires modeling
  $\mathbf{R}_i$ directly rather than inferring it indirectly from how well a random-slope
  model happens to fit.

And, as before, there's a structural reason you can't simply default to `nlme` for everything:
`nlme` handles nested random effects gracefully but cannot fit fully **crossed** random effects
at all (Section 8, participants crossed with items/stimuli), while `lme4` handles crossed
random effects easily but cannot model residual correlation structures like `corAR1`. In
practice: reach for `nlme` when you specifically need $\mathbf{R}$-side structure for one of
the four reasons above; reach for `lme4` (with as rich a $\mathbf{G}$ as your design supports)
for crossed sampling units, and trust it more than you might expect for fixed-effect inference
even with longitudinal dependency. If you need crossed random effects *and* explicit
$\mathbf{R}$-side structure at once, `glmmTMB` (Section 7) is increasingly a tool of choice,
since it supports both within the same model.

## Summary

* $\mathbf{G}$ (random effects) and $\mathbf{R}_i$ (residuals) both contribute to
  $\text{Var}(\mathbf{y}_i)$, and a sufficiently rich $\mathbf{G}$ -- like the random-slope
  growth-curve models built in Sections 2 and 4 -- can already capture a meaningful share of
  real autocorrelation, without ever touching $\mathbf{R}_i$. `lme4`'s inability to model
  $\mathbf{R}_i$ directly is a real but narrow limitation, not a general inadequacy for
  longitudinal data.
* Fixed-effect point estimates are robust to how dependency is partitioned between
  $\mathbf{G}$ and $\mathbf{R}_i$. Fixed-effect standard errors are *not* automatically robust
  to this -- they tend to be well-calibrated when $\mathbf{G}$ is rich enough to approximate
  the true dependency (e.g., a random slope that can absorb "streaky," time-correlated
  patterns), and can be meaningfully anti-conservative when $\mathbf{G}$ is sparse (e.g., if you only have 
  a random intercept, which cannot represent anything beyond a constant, compound-symmetric
  correlation).
* The variance components themselves are considerably more sensitive to this partition than
  the fixed effects are -- a model can pass a fixed-effect calibration check while still
  giving a biased answer to "how much do subjects differ," because it achieved that
  calibration by misattributing dependency to the wrong variance component.
* `nlme::lme()` lets you specify an explicit residual correlation structure
  (`correlation = corAR1(...)`, `corCAR1(...)`, `corCompSymm(...)`, `corSymm(...)`) and/or an
  explicit variance function (`weights = varIdent(...)`, among others) -- reach for it when
  the variance components are your actual question, your time structure doesn't fit a
  low-order polynomial random-effects structure, your $\mathbf{G}$ is sparse relative to the
  true dependency, or the dependency structure is itself the finding.
* Compare models sequentially with `anova()` (a nested-model likelihood-ratio test, just as
  we did for the unconditional growth models in Section 2) rather than jumping straight to
  the most complex structure.
* Misspecifying one piece of the covariance structure (e.g., leaving out heteroscedasticity)
  can bias your estimates of a different piece (e.g., the autocorrelation parameter) --
  inspect more than one diagnostic before concluding a model is "good enough."


## References
Pinheiro, J., & Bates, D. (2006). *Mixed-effects models in S and S-PLUS*. Springer.

Pinheiro, J., Bates, D., & R Core Team. (2022). nlme: Linear and Nonlinear Mixed Effects
Models. https://CRAN.R-project.org/package=nlme
