---
title: 'Section 3: Mixed-Effects Models for Factorial Designs'
author: "Keith Lohse, PhD, PStat"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

## Note that this page is in the process of being updated 2022-10-04

For these examples, we are going to work with a fictional data that that has two 
fully nested (i.e., between-subjects) and two fully crossed (i.e., within-subject)
factors. We have two hypothetical groups of older adults (OA, >65 y) and younger 
adults (YA, <35 y). Within each of those groups, half of the participants were
in the control group (C) and half were in the treatment group (T). Each participant 
was also measured on four different trials (1 to 4) in three different conditions 
(A, B, and C). 

First, we will open the relevant R packages that we will use and then we will 
download the data from the web. (If you don't have these packages installed, you 
will need install the packages before running the library() function.) Once the data 
are downloaded we can use the head() function to view the first ten rows of the data.


```{r, setting libraries, results="hide", message = FALSE, warning=FALSE}
library(tidyverse); library(RCurl); library(ez); library(lme4); library(car); library(lmerTest)

DATA <- read.csv("https://raw.githubusercontent.com/keithlohse/mixed_effects_models/master/data_AGING_example.csv",
                 stringsAsFactors = TRUE)

```

``` {r, print data to screen}
head(DATA, 10)

```

For the examples we will walk through, we will ignore different variables at different 
times. For instance, we will want to average across trials to create a data set with one 
observation per condition (which we will call *data_COND*), and we will want to average across conditions
to create a data set with one observation for each trial (which we will call *data_TIME*).

Averaging across trials, here are the first ten rows of *data_COND*. 
```{r, subsetting data by condition, warning=FALSE}
data_COND <- DATA %>% group_by(subID, condition, age_group, group) %>% # Note we ignore trial to average across it
  summarize(speed = mean(speed, na.rm=TRUE)) %>% # I have included na.rm=TRUE even though there are no missing data
  arrange(age_group, subID, condition) # finally, we can resort our data by subIDs within groups
head(data_COND, 10)
```

Averaging across conditions, here are the first ten rows of *data_TIME*. 
``` {r, subsetting data by time, warning=FALSE}
data_TIME <- DATA %>% group_by(subID, time, age_group, group) %>% # Note we ignore condition to average across it
  summarize(speed = mean(speed, na.rm=TRUE)) %>% # I have included na.rm=TRUE even though there are no missing data
  arrange(age_group, subID, time) # finally, we can resort our data by subIDs within groups
head(data_TIME, 10)

```

With these data in place, we can now test all of our different models. We are going
to explore the appropriate mixed-effects regression (MER) models for these different 
situations:

  1. **MER as One-Way Repeated Measures ANOVA**
      * When we have a single crossed factor (i.e., one within-subjects factor) in 
      our mixed-effects model.
  2. **MER as Two-Way Repeated Measures ANOVA**
      * When we have multiple crossed factors (i.e., two within-subjects factors) in 
    our mixed-effects model.
  3. **MER as Mixed-Factorial ANOVA with a Single Within-Subjects Factor**
      * When we have a single crossed factor (i.e., one within-subjects factor) and 
    a single nested factor (i.e., between-subjects factors) in our mixed-effects 
    model.
  4. **MER as Mixed-Factorial ANOVA with Multiple Within-Subjects Factors**
      * When we have multiple crossed factors (i.e., two within-subjects factor) and 
    a single nested factor (i.e., between-subjects factors) in our mixed-effects 
    model.
  5. **MER as Three-Way Repeated-Measures ANOVA (No Between-Subjects Factor)**
      * When we have three crossed factors (i.e., three within-subjects factors) and
    *no* nested, between-subjects factor at all in our mixed-effects model. This
    situation is a natural extension of Sections 1 and 2 above (which handled one
    and two crossed factors, respectively) and it is also exactly the situation
    that Lohse, Kozlowski, and Strube (2023; the companion paper to this repository)
    describe when they discuss adding a third within-subject factor to a fully
    factorial design. We work through a fully-verified example below, including
    what happens when the "textbook correct" random-effects structure runs into
    trouble.


# 1. MER as One-Way Repeated Measures ANOVA 
## (A Single Crossed Factor)
For this example, we will focus on only the effect of condition, so we will use 
the data_COND dataset to average across different trials. First, let's plot the 
data to get a better sense of what the data look like. 


``` {r plotting the effects of condition, echo=TRUE, fig.align="center"}
ggplot(data_COND, aes(x = condition, y = speed)) +
  geom_point(aes(fill=condition), pch=21, size=2,
             position=position_jitter(w=0.2, h=0))+
  geom_boxplot(aes(fill=condition), col="black", 
               alpha=0.4, width=0.5, outlier.shape = NA)+
  scale_x_discrete(name = "Condition") +
  scale_y_continuous(name = "Speed (m/s)") +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "none")
  
```

## 1.1. As an ANOVA...
To implement a simple one-way repeated measures ANOVA, we have a few options. We 
could directly code our ANOVA using the aov() function in R:
``` {r one way ANOVA the hard way, echo=TRUE, warning=FALSE}
summary(aov(speed ~ condition + Error(subID/condition), data=data_COND))
```

Or we can use the ezANOVA() function from the "ez" package. 
```{r one way ANOVA the ez way, echo=TRUE, warning=FALSE}
ezANOVA(data = data_COND, 
    dv = .(speed),
    wid = .(subID),
    within = .(condition)
)
```
Although these two different functions present the results in slightly different
ways, note that the f-values for the two different omnibus tests match. By either 
method, the f-observed for the main-effect of condition is **F(2,78)=24.46, p<0.001**. 

If you are familiar with issues of contrast versus treatment coding, Type I
versus Type III Sums of Squared Errors, and collinearity, then directly controlling
your own ANOVA using base R is probably a safe bet. 

If you are not familiar with those terms/issues, then the ezANOVA code is probably the best option for you. However, I would strongly encourage you to develop a more detailed understanding of general-linear models before jumping into mixed-effect models. 

By default the aov() function provides a test of your statistical
model using Type I Sums of Squared Errors, whereas the ezANOVA() function provides
a test of your statistical model using Type III Sums of Squared Errors. In a completely
orthogonal design where all factors are statistically independent of each other 
(i.e., there is no collinearity), the Type I and Type III Sums of Squared Errors 
will agree. By default, R also uses treatment coding (or "dummy" coding) for categorical
factors rather than orthogonal contrast codes. The omnibus F-tests that we see in 
the ANOVA output will be the same regardless of the types of codes used, but if 
we dig into individual regression coefficients, it is important to remember how
these variables were coded so that we can interpret them correctly.


## 1.2. Getting the same result with a mixed-effect model...
Because we have a single within-subject factor, we will  need to add a 
random-effect of subject to account for individual differences between subjects.
By partitioning the between-subjects variance out of our model, we can fairly test
the effect of *condition*, because our residuals will now be independent of each
other. 
```{r one way ANOVA in a mixed effects model, echo=TRUE, warning=FALSE}
# First we will define our model
mod1 <- lmer(speed ~ 
               # Fixed Effects:
               condition + 
               # Random Effects: 
               (1|subID), 
             # Define the data: 
             data=data_COND, REML = TRUE)

# We can then get the ANOVA results for our model:
anova(mod1)
```

Most critically, note that the F-value, **F(2,78)=24.46, p<0.001** is identical in 
both the RM ANOVA and in the mixed-effects regression model.

If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects
and residuals. 

```{r}
summary(mod1)
```







# 2. MER as Two-Way Repeated Measures ANOVA 
## (Multiple Crossed Factors)
For this example, we will be analyzing both Time and Condition, so we will use 
our full data frame, *DATA*. First, let's plot the data to get a better 
sense of what our data look like.

``` {r plotting two repeated measures, fig.align="center"}
DATA$time <- factor(DATA$time)

ggplot(DATA, aes(x = time, y = speed)) +
  geom_point(aes(fill=condition), pch=21, size=2,
             position=position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1))+
  geom_boxplot(aes(fill=condition), col="black", 
               alpha=0.4, width=0.5, outlier.shape = NA)+
  labs(fill="Condition")+
  scale_x_discrete(name = "Trial") +
  scale_y_continuous(name = "Speed (m/s)", limits = c(0,3)) +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16, color="black"), 
        legend.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "bottom")
``` 


## 2.1. As an ANOVA...
Before we run our models, we want to convert trial to a **factor** so that our model
is treating time categorically rather than continuously. (In later modules, we 
will discuss how to mix continuous and categorical factors.) Additionally, remember
that we are now using our larger data set *DATA* rather than our aggregated data
set *data_COND*. To implement a two-way repeated measures ANOVA, we have the same 
options as before. We can directly code our ANOVA using the aov() function in R:
``` {r two way ANOVA the hard way, echo=TRUE}
DATA$time <- factor(DATA$time)
summary(aov(speed ~ condition*time + Error(subID/(condition*time)), data=DATA))
```

Or we can use the ezANOVA() function from the "ez" package. 
```{r two way ANOVA the ez way, echo=TRUE}
ezANOVA(data = DATA, 
    dv = .(speed),
    wid = .(subID),
    within = .(time, condition)
)
```
Again, regardless of the coding approach that you use, these two different functions 
produce the same f-observed for the main-effects of time, **F(3,117)=105.95, p<0.001**,
and Condition, **F(2,78)=24.46, p<0.001**, and the Time x Condition interaction, **F(6,234)=13.53, p<0.001**. 

 


## 2.2. Two-way RM ANOVA as a mixed-effect model...
Because we now have two crossed factors, we need to account not only for the fact 
that we have multiple observations for each participant, but we have multiple 
observations for each person at each level of each factor. 

For instance, for the effect of Condition, we actually have Condition effects at Trial 1, Trial 2, Trial 3, and Trial 4. 
The converse is also true for the effect of Time, we have four different trials in
Condition A, Condition B, and Condition C. Thus, in order to appropriately 
account for the statistical dependencies in our data, we need to add random-effects
of "time:subID" and "condition:subID" to the model. 

The colon operator (":") means that we are crossing or multiplying these factors. 
That is, if we have subject ID's A, B, and C and Trials 1, 2, 3, and 4, then we end up with 
A1, A2, A3, A4, B1, B2, B3, B4, etc. 

For more information on how random-effects are specified and what they mean in R,
I recommend looking at this discussion on Stack Exchange: https://stats.stackexchange.com/questions/228800/crossed-vs-nested-random-effects-how-do-they-differ-and-how-are-they-specified 


```{r two way another in a mixed effects model, echo=TRUE, warning=FALSE}
# First we will define our model
mod2 <- lmer(speed ~ 
               # Fixed Effects:
               time*condition + 
               # Random Effects
               (1|subID)+ (1|time:subID) + (1|condition:subID), 
               # Define your data, 
             data=DATA, REML=TRUE)

# We can then get the ANOVA results for our model:
anova(mod2)
```

Again, for our purposes, the critical thing to note is that the F-values for the 
main-effects and interactions are the same between our RM ANOVAs and the mixed-effects
model. Without delving into the mathematical details, this is a good demonstration 
that the appropriate random-effects our regression model make it analogous to the
factorial ANOVA. This allows us to capitalize on the benefits of mixed-effects regression for
designs that we would normally analyze using factorial ANOVA. 


If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects
and residuals. 

```{r}
summary(mod2)
```





# 3. MER as Mixed-Factorial ANOVA with One Crossed Factor 
## 2 (Age) x 3 (Condition) ANOVA Example
For this example, we will now consider both the between-subject factor of age, 
as well as the within-subject factor of condition. Do to this, we will 
average over the different trials (1, 2, 3, and 4) to get one observation in each 
condition. This design is sometimes referred as a "mixed-factorial" design, because
we have a mix of between-subjects and within-subject factors. Depending 
on your background, you might be more familiar with this as a "split plot" design. 
Split plot comes from agronomy (where a lot of statistics where developed!) and 
refers to the fact that some plots are assigned to different levels of Factor A (e.g., Treatment
versus Control), but within each plot there is a second level of randomization to 
different levels of Factor B (e.g., Conditions A, B, or C). Both terms describe 
the same thing but differ in their unit of analysis. In psychology, a single person 
is usually the experimental unit (hence "within-subject" variables) whereas in
agronomy or biology a physical area might be the unit of analysis (hence "split-plot"
variables). The more general way to talk about these terms is as **nested-** versus 
**crossed-**factors. 

* For fully **nested** factors, an experimental unit is represented in only
one of the factors (e.g., a person can only be in the treatment group or the 
control group). 
* For fully **crossed** factors, an experimental unit is represented at
all levels of the factor (e.g., e.g., a person is tested in conditions A, B, and C).

In our example, participants are nested within age-group, because each person
is represented at only one level of that factor (i.e., you are only a younger adult or older adult). 
However, Condition is a crossed factor, because each person is represented at all levels of Condition 
(i.e., each person was measured in all three conditions). We can see this more clearly if we plot all of our data.

``` {r plotting crossed and nested factors, warning=FALSE, fig.align="center"}
ggplot(data_COND, aes(x = condition, y = speed)) +
  geom_point(aes(fill=age_group), pch=21, size=2,
             position=position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1))+
  geom_boxplot(aes(fill=age_group), col="black",
               alpha=0.4, width=0.5, outlier.shape=NA)+
  labs(fill="Age Group")+
  scale_x_discrete(name = "Condition") +
  scale_y_continuous(name = "Speed (m/s)") +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16, color="black"), 
        legend.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "bottom")
``` 


## 3.1. As an ANOVA...
For this mixed factorial ANOVA, we have one factor of Condition that varies 
within-subjects, but we also have a factor of Age Group that 
varies between subjects. As before, we can directly code this into our analysis
of variance using the aov() function or using the ezANOVA() function from the "ez" 
package. 
``` {r mixed factorial ANOVA the hard way, echo=TRUE}
summary(aov(speed ~ age_group*condition + Error(subID/condition), data=data_COND))
```

Or we can use the ezANOVA() function from the "ez" package. 
```{r mixed factorial ANOVA the ez way, echo=TRUE}
ezANOVA(data = data_COND, 
    dv = .(speed),
    wid = .(subID),
    within = .(condition),
    between = .(age_group)
)
```

The ezANOVA() function provides us with more detail, for instance automatically
providing Mauchly's Test for Sphericity and both the Greenhouse-Geisser and Hyunh-Feldt
corrections in the event that the sphericity assumption is violated. Most importantly, however,
the outputs of these two functions agree when we look at the f-statistics for all
main-effects and interactions when sphericity is assumed.

 


## 3.2. Mixed Factorial ANOVA as a mixed-effect model...
For this mixed-factorial design, we need to account for the fact that we have
multiple observations coming from each person, so we will add a random-effect of 
"subID". After accounting for this statistical dependence in our data, we can now 
fairly test the effects of Age Group, and Condition with residuals that are
independent of each other. 

```{r mixed factorial ANOVA in a mixed effects model, echo=TRUE}
# First we will define our model
mod3 <- lmer(speed ~ 
               # Fixed Effects:
               age_group*condition + 
               # Random Effects
               (1|subID), 
               # Define your data, 
             data=data_COND, REML=TRUE)

# We can then get the ANOVA results for our model:
anova(mod3)
```

Critically, note that the f-statistics and degrees of freedom all match our ANOVA outputs (when sphericity is assumed), 
with a main-effect of Age Group, F(1,38)=47.18, Condition, F(2,76)=27.27, and the 
Age Group x Condition interaction, F(2,76)=5.48.

If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects,
and residuals. 

```{r}
summary(mod3)
```



# 4. MER as Mixed-Factorial ANOVA with Multiple Crossed Factors 
## 2 (Age Group) x 3 (Condition) x 4 (Trial) ANOVA
Next, we will consider the more complicated example of our fully factorial design
with a nested factor of Age-Group, and crossed factors of Condition and
Time.

``` {r plotting multiple crossed and nested factors, fig.align="center"}
DATA$time <- factor(DATA$time)

ggplot(DATA, aes(x = time, y = speed)) +
  geom_point(aes(fill=age_group), size=2, shape=21,
             position=position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1))+
  geom_boxplot(aes(fill=age_group), col="black", 
               alpha=0.4, width=0.5, outlier.shape = NA)+
  labs(fill="Age Group")+
  facet_wrap(~condition)+
  scale_x_discrete(name = "Trial") +
  scale_y_continuous(name = "Speed (m/s)", limits = c(0,3)) +
  theme(axis.text=element_text(size=16, color="black"), 
        axis.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16, color="black"), 
        legend.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=16, face="bold", hjust=0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=16, face="bold"),
        legend.position = "bottom")
``` 


## 4.1. As an ANOVA...
For this mixed factorial ANOVA, we have two factors that vary within subjects, 
Condition and Time, and we have one factor that varies between subjects, 
Age Group. As before, we can directly code this into our analysis
of variance using the aov() function or using the ezANOVA() function from the "ez" 
package. 
``` {r multiway factorial ANOVA the hard way, echo=TRUE, warning=FALSE}
summary(aov(speed ~ age_group*condition*time + Error(subID/(condition*time)), data=DATA))
```

```{r multiway factorial ANOVA the ez way, echo=TRUE, warning=FALSE}
ezANOVA(data = DATA, 
    dv = .(speed),
    wid = .(subID),
    within = .(condition, time),
    between = .(age_group)
)
```



## 4.2. Multi-Way Mixed Factorial ANOVA as a mixed-effect model...
For this mixed-factorial design, we need to account for the fact that we have
multiple observations coming from each person, but we also need to acocunt for the 
fact that we multiple observations for each Condition and on each Trial. In order
to account for this dependence in our data, we need to include random-effects
of subject, subject:condition, and subject:time. Adding these random-effects to our 
model will make our mixed-effects model statistically equivalent to the mixed-factorial
ANOVAs that we ran above.

```{r multiway factorial ANOVA in a mixed effects model, echo=TRUE}
# First we will define our model
mod4 <- lmer(speed ~ 
               # Fixed Effects:
               age_group*condition*time + 
               # Random Effects
               (1|subID)+(1|condition:subID)+(1|time:subID), 
               # Define your data, 
             data=DATA, REML=TRUE)

# We can then get the ANOVA results for our model:
anova(mod4)
```

Note that the f-statistics and the degrees of freedom match what we got from the 
ANOVA tables above (when sphericity is assumed). There are a lot of effects so I 
wont list them all, but note the main-effects of Age Group, F(1,38)=47.18, 
Condition, F(2,76)=27.27, and Time, F(3,114)=113.88.

Readers familiar with repeated measures ANOVA wont be surprised to see that that the 
output of the aov() function partitions the effects into different sections. The 
between-subject effect of Age Group is evaluated using residual error at the level of 
"Error:subID". The effects of Condition and Age Group x Condition are evaluated 
using residuals of "Error: subID:condition". This is precisely what we are doing
when we include the three different random-intercepts in the mixed-effects regression.
The $(1|subID)$ partitions out the average individual differences between subjects, 
$(1|condition:subID)$ and $(1|time:subID)$ partitions out the variance at the different
levels of our repeated measures.

If we want to delve deeper into our model, we can also use the summary() function
to get more information about model fit statistics, parameter estimates, random-effects
and residuals. 

```{r}
summary(mod4)
```

# 5. MER as Three-Way Repeated-Measures ANOVA 
## (Three Crossed Factors, No Between-Subjects Factor)
In Sections 1 and 2 above, we built up the mixed-effects analogue of a repeated-measures
ANOVA one crossed factor at a time (one-way, then two-way). In Sections 3 and 4, we then
added a *between-subjects* factor (Age Group) alongside one or two *within-subjects*
factors. But we have not yet pushed the purely within-subjects case (no between-subjects
factor at all) beyond two crossed factors. This is worth doing for its own sake, because a
fully within-subjects, three-way design is common in rehabilitation and movement science
(e.g., every participant experiences every combination of Load x Terrain x Fatigue, or
every combination of Medication x Time-of-Day x Task), and because it is exactly the
situation that our companion paper (Lohse, Kozlowski, & Strube, 2023, *Communications in
Kinesiology*) flags as getting "more complicated" once a third within-subject factor is
added, but only sketches in a single sentence and footnote. Below, we build out a fully
worked, numerically-verified example, including a very important complication that shows
up in real (not just hypothetical) data: what happens when the "textbook correct"
random-effects structure runs into estimation trouble.

For this example, we simulate a new hypothetical data set (`data_GAIT_3way_example.csv`)
in which 32 participants walked an obstacle course under every combination of three fully
crossed within-subject factors:

  * **Load**: carrying a **Light** vs. a **Heavy** pack.
  * **Terrain**: walking on **Flat** ground, **Uneven** ground, or over **Stairs**.
  * **Fatigue**: tested **Pre**- vs. **Post**-fatigue protocol.

Every participant is measured exactly once under all $2 \times 3 \times 2 = 12$
combinations of these factors (384 observations total), and there is no between-subjects
factor at all (e.g., no separate age or treatment groups). The R script used to simulate
this data set (`simulate_data_GAIT_3way.R`) is included in the repository alongside the
data, so that instructors can regenerate it, inspect the true, known effects that were
simulated, or adapt the script to build their own examples with different numbers of
factors, levels, or subjects.

```{r, reading in 3way within-subjects data, results="hide", message = FALSE, echo=FALSE}
DAT3 <- read.csv("https://raw.githubusercontent.com/keithlohse/mixed_effects_models/master/data_GAIT_3way_example.csv",
                 stringsAsFactors = TRUE)
DAT3$load    <- factor(DAT3$load,    levels = c("Light","Heavy"))
DAT3$terrain <- factor(DAT3$terrain, levels = c("Flat","Uneven","Stairs"))
DAT3$fatigue <- factor(DAT3$fatigue, levels = c("Pre","Post"))
```

```{r plotting 3way within subjects data, fig.align="center", warning=FALSE}
ggplot(DAT3, aes(x = terrain, y = speed, fill = load)) +
  geom_boxplot(col="black", alpha=0.6, width=0.6, outlier.shape = NA,
               position = position_dodge(width=0.7)) +
  geom_point(pch=21, size=1.5, alpha=0.6,
             position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.1)) +
  facet_wrap(~fatigue) +
  labs(fill="Load") +
  scale_x_discrete(name = "Terrain") +
  scale_y_continuous(name = "Speed (m/s)") +
  theme(axis.text=element_text(size=14, color="black"), 
        axis.title=element_text(size=14, face="bold"),
        legend.text=element_text(size=14, color="black"), 
        legend.title=element_text(size=14, face="bold"),
        strip.text = element_text(size=14, face="bold"),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")
```

## 5.1. As an ANOVA...
Because *every* factor here is within-subjects, we account for the statistical dependence
of every observation coming from the same person by nesting the *entire* factorial design
under subject in our `Error()` term:
```{r three way RM ANOVA the hard way, echo=TRUE, warning=FALSE}
aov_3way <- aov(speed ~ load*terrain*fatigue + Error(subID/(load*terrain*fatigue)), data=DAT3)
summary(aov_3way)
```
Notice that this single `Error()` call produces **eight** different strata: the (unused,
since we have no between-subjects factor) `Error: subID` stratum, one stratum for each of
the three main effects (`subID:load`, `subID:terrain`, `subID:fatigue`), one stratum for
each of the three two-way interactions (`subID:load:terrain`, `subID:load:fatigue`,
`subID:terrain:fatigue`), and finally the three-way interaction stratum
(`subID:load:terrain:fatigue`). Because each subject contributes exactly one observation
to every cell of the design, this final three-way stratum is also serving as our "leftover"
residual error -- there is no deeper partition of the data left to make.

## 5.2. Three-Way RM ANOVA as a mixed-effect model...
To reproduce these eight strata in a single `lmer()` call, we can extend the same logic we
used for the two-way, purely within-subjects case in Section 2 above. There, to account for
two crossed factors we needed random intercepts for `subject`, `subject:A`, and `subject:B`,
and the model's own residual automatically played the role of the finest (`subject:A:B`)
stratum. For **three** crossed factors, we push that logic one level deeper: we need a
random intercept for `subject`, one for each **main effect nested in subject**
(`subject:load`, `subject:terrain`, `subject:fatigue`), and one for each **two-way
interaction nested in subject** (`subject:load:terrain`, `subject:load:fatigue`,
`subject:terrain:fatigue`). The model's residual then automatically absorbs the finest,
three-way stratum (`subject:load:terrain:fatigue`), which is exactly the stratum used to
test the three-way interaction in the classical ANOVA above.

```{r three way in a mixed effects model, echo=TRUE, warning=FALSE}
# First we will define our model
mod5 <- lmer(speed ~ 
               # Fixed Effects:
               load*terrain*fatigue + 
               # Random Effects: one term for subject, one for each factor 
               # nested in subject, and one for each two-way interaction 
               # nested in subject:
               (1|subID) + 
               (1|subID:load) + (1|subID:terrain) + (1|subID:fatigue) +
               (1|subID:load:terrain) + (1|subID:load:fatigue) + (1|subID:terrain:fatigue), 
             data=DAT3, REML=TRUE)

# We can then get the ANOVA results for our model:
anova(mod5)

# It is always worth checking whether the model actually converged cleanly:
isSingular(mod5)
```
With `isSingular(mod5)` returning `FALSE`, this model converged without hitting a
boundary, and -- as promised -- every F-value and every denominator degrees of freedom
matches the classical repeated-measures ANOVA above exactly: Load, F(1,31)=375.52;
Terrain, F(2,62)=207.06; Fatigue, F(1,31)=122.77; Load x Terrain, F(2,62)=15.76; Load x
Fatigue, F(1,31)=14.89; Terrain x Fatigue, F(2,62)=0.14; and Load x Terrain x Fatigue,
F(2,62)=0.21. As in Sections 1, 2, and 4, this is a good demonstration that the
appropriately-specified mixed-effects model is statistically equivalent to the classical
repeated-measures ANOVA for a complete, balanced, fully-crossed design.

**A general recipe.** Generalizing across sections: for $k$ fully crossed,
purely within-subjects factors with no missing cells, you need a random intercept of
`subject`, plus one random intercept for `subject` crossed with every combination of 1 up
to $k-1$ of your within-subjects factors. The model's residual then automatically plays
the role of the finest, $k$-way stratum. For $k=1$: `(1|subject)`. For $k=2$:
`(1|subject) + (1|subject:A) + (1|subject:B)`. For $k=3$: the seven-term structure used
in `mod5` above. Notice how quickly this grows -- a fourth crossed factor would require
$1 + 4 + 6 = 11$ random-effect terms before the residual even gets involved! This rapid
growth in complexity is itself an important, practical reason to think carefully before
adding more and more crossed within-subject factors to a single design (see Section 5.3
below).

## 5.3. A Note on Boundary (Singular) Fits in Saturated, Three-Way Designs
The random-effects structure above is the theoretically "correct" one, in the sense that
it reproduces the classical ANOVA's error strata exactly. But it asks REML to estimate
*seven* separate variance components from data that, ultimately, come from only 32
subjects -- and some of those variance components (like `subject:load:fatigue`, which is
the interaction of two factors that each only have **two** levels) reflect only a single
contrast per subject. This is a very easy structure to over-ask of your data, and it is
worth seeing exactly what happens when you do.

To illustrate, let's refit the identical model, but using only the first 12 (of 32)
subjects from the very same simulated population:
```{r singular fit demonstration, echo=TRUE, warning=TRUE}
DAT3_small <- DAT3 %>% filter(subID %in% levels(subID)[1:12]) %>% droplevels()

mod5_small <- lmer(speed ~ 
                      load*terrain*fatigue + 
                      (1|subID) + 
                      (1|subID:load) + (1|subID:terrain) + (1|subID:fatigue) +
                      (1|subID:load:terrain) + (1|subID:load:fatigue) + (1|subID:terrain:fatigue), 
                    data=DAT3_small, REML=TRUE)

isSingular(mod5_small)
VarCorr(mod5_small)
anova(mod5_small)
```
With only 12 subjects, R warns us of a "boundary (singular) fit," and looking at
`VarCorr(mod5_small)` shows why: the variance estimate for `subID:load` has been pushed
all the way down to 0.00. Load only has two levels, so `subject:load` reflects a single
per-subject contrast -- exactly the kind of variance component that is hardest to
distinguish from residual noise with a modest sample size.

Critically, this is not just a cosmetic warning. Because `subID:load` collapsed into the
residual, **every fixed effect that depends on that stratum** now uses a different,
non-integer denominator degrees of freedom, and no longer matches the classical ANOVA
exactly: Load itself (F=160.32, df=1,21.1, versus the ANOVA's F=188.2, df=1,11), Load x
Terrain (F=3.08, df=2,26.1, versus the ANOVA's F=2.94, df=2,22 -- note this even flips from
non-significant to a different p-value), and Load x Fatigue (F=8.93, df=1,17.9, versus the
ANOVA's F=7.97, df=1,11) have all drifted noticeably from their ANOVA counterparts. Terrain,
Fatigue, and Terrain x Fatigue -- none of which depended on the collapsed `subID:load` term
-- still match almost exactly (e.g., Terrain: F=108.75 vs. the ANOVA's 108.7, both with
df=2,22). The three-way interaction shows only a small drift (F=0.40, df=2,21.6, versus the
ANOVA's F=0.41, df=2,22), since it is one step further removed from the collapsed term. In
other words, a singular fit does not just add a scary warning message; in a design like
this one, it can measurably erode the very ANOVA-equivalence property that was the whole
reason for building this random-effects structure in the first place, and the terms most
directly tied to the collapsed variance component are affected the most.

This is precisely the tension that Barr, Levy, Scheepers, & Tily (2013) and Bates, Kliegl,
Vasishth, & Baayen (2015) debate under the labels "keep it maximal" versus "parsimonious
mixed models" (the latter, along with a closely related paper by Barr alone, is already in
the reference list of the companion *Communications in Kinesiology* paper for this
repository), and that Matuschek, Kliegl, Vasishth, Baayen, & Bates (2017) resolve
empirically: keeping every theoretically-justified
random effect can protect against inflated Type I error, but it can also cost real
statistical power once a variance component is not well-supported by the data, and (as
shown here) it does not even guarantee the exact ANOVA-equivalence it was meant to
provide. In practice, we would suggest:

1. **If your design is complete and balanced** (as it is here -- every subject has exactly
   one observation in every cell, with no missing data), the classical repeated-measures
   ANOVA (`aov()` or `ezANOVA()`) will always give you the "gold standard" answer directly,
   without any risk of a singular fit. The main reasons to reach for the mixed-effects
   version of a fully-crossed, purely within-subjects design are (a) you *do* have some
   missing data or an unbalanced design, where the classical ANOVA's `Error()` syntax
   breaks down but `lmer()` still works, or (b) you specifically want to report and
   interpret the variance components themselves (e.g., "how much do people vary in their
   Load x Fatigue effect?").
2. **If you need the mixed-effects version**, consider a more parsimonious structure:
   dropping the correlations among random effects with the `||` operator, or dropping the
   specific variance components that are estimated at (or very near) the zero boundary,
   and refitting. This will not recover the *exact* ANOVA F-values, but it can resolve the
   convergence warning and typically improves statistical power relative to the maximal
   model.
3. **Plan for this ahead of time.** If you know your design will have three or more fully
   crossed within-subjects factors, simulating data first (as we did to build this section)
   is a cheap way to check whether your planned sample size can support a maximal
   random-effects structure at all, before you collect any real data.


## References for this Section
Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language, 68*(3), 255-278. https://doi.org/10.1016/j.jml.2012.11.001

Bates, D., Kliegl, R., Vasishth, S., & Baayen, H. (2015). Parsimonious Mixed Models (Version 2). *arXiv*. https://doi.org/10.48550/ARXIV.1506.04967

Lohse, K. R., Kozlowski, A. J., & Strube, M. J. (2023). Model Specification in Mixed-Effects Models: A Focus on Random Effects. *Communications in Kinesiology*. https://doi.org/10.51224/cik.2023.52

Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language, 94*, 305-315. https://doi.org/10.1016/j.jml.2017.01.001
